### 2012-12-11 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updated mandc paper for typos

%\usepackage{epsf}
%
%
% Insert authors' names and short version of title in lines below
%
\newcommand{\authorHead}      % Author's names here
   {Slattery, Wilson, and Evans}  
\newcommand{\shortTitle}      % Short title here
   {Domain Decomposed Neumann-Ulam Method}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%   BEGIN DOCUMENT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\section*{ABSTRACT} 
\begin{quote}
\begin{small}
The domain decomposed behavior of the adjoint Neumann-Ulam Monte Carlo
method for solving linear operator equations is analyzed using the
spectral properties of the linear operator. Relationships for the
average length of the adjoint random walks, a measure of convergence
speed and serial performance, are made with respect to the Eigenvalues
and other properties of the linear operator. In addition,
relationships for the effective optical thickness of a domain in the
decomposition are presented based on the spectral analysis and
diffusion theory. Using the effective optical thickness, the Wigner
rational approximation and the mean chord approximation are applied to
estimate the leakage fraction of stochastic histories from a domain in
the decomposition as a measure of parallel performance and potential
communication costs. The one-speed, two-dimensional neutron diffusion
equation is used as a model problem to test the models for symmetric
operators. In general, the derived approximations show good agreement
with measured computational results.

\emph{Key Words}: Monte Carlo, domain decomposition, MCSA, parallel
computing
\end{small} 
\end{quote}

@@

To enable higher fidelity reactor physics Monte Carlo simulations,
domain decomposition has been identified as a key principle in moving
forward in high performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because these nearest-neighbor
communication sequences have poor algorithmic strong scaling
\cite{gropp_high-performance_2001}.

The purpose of this study is to provide a simple, analytic theory
based on the properties of the linear system the will allow for
estimates of the domain decomposed behavior of the adjoint
Neumann-Ulam method. When solving problems where the linear operator
is symmetric, a host of analytic theories exist based on the
Eigenvalue spectrum of the operator that characterize their behavior
in the context of deterministic linear solvers. Using past work, these
theories are adapted to the domain decomposed adjoint Neumann-Ulam
method using the one-speed, two-dimensional neutron diffusion
equation. In this paper we describe the adjoint Neumann-Ulam Monte
Carlo method followed by a presentation of the model problem. Using
the linear system generated by this discretization, we use a spectral
theory to generate analytic relations for the Eigenvalues of the
operator based on system parameters. Using the Eigenvalue spectra, we
then build relationships to characterize the transport of stochastic
histories in a decomposed domain and the fraction of histories that
leak from a domain and will therefore have to be
communicated. Finally, we compare these analytic results to numerical
experiments conducted with the model problem and draw conclusions
looking towards future work.

@@

In order to perform the random walks required to approximate the
summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build
transition probability and weight matrices where probabilities are
column-scaled and the transition weight is defined as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
  \frac{h_{ji}}{p_{ij}},
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $i$ from the current state
in the random walk $j$ by sampling column-wise.  Using our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the contribution to the solution in state $i$
from a particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ will be the sampled source starting
weight. Finally, the expectation value using all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimate of the
solution.

@@

The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the Eigenvalues of the iteration matrix. We
will compute these Eigenvalues by assuming Eigenfunctions of the form
\cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
Eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these Eigenfunctions should
give the Eigenvalues of the matrix as they lie on the unit circle in
the complex plane.

@@

The spectral radius of the iteration matrix is obtained by seeking its
largest Eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the Eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the Eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. We expect the maximum Eigenvalue to exist when $p=q=0$, giving
the following for the spectral radius of the Jacobi preconditioned
iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

@@

In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\cite{dimov_new_1998,halton_sequential_1994,danilov_asymptotic_2000}. This
random walk length, or the number of transitions before the
termination of a history (either by the weight cutoff, absorption, or
exiting the domain) is therefore approximately the number of
stationary iterations required to converge to the specified
tolerance. In the case of the adjoint Neumann-Ulam method, no such
tolerance exists, however, we have specified a weight cutoff $W_c$
under which low-weight histories will be prematurely terminated as
their contributions are deemed minute. After $k$ iterations, a
stationary method is terminated as the error has reached some
fraction, $\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
Eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

@@

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a domain in the
decomposition and the corresponding leakage fraction in terms of
properties of the linear operator and the discretization. To begin we
must first calculate the mean distance a Monte Carlo history will move
in the grid by computing the mean squared distance of its movement
along the chord defined across the domain of length $l$. After a
single transition a history will have moved a mean squared distance
of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the discrete grid elements along the chord
and $n_s$ is the number of grid elements a history will move on
average every transition. For our diffusion model problem, $n_s$ would
equate to the number of states in the $i$ (or $j$ as the problem is
symmetric) direction that a history could potentially move in a single
transition and is dependent on the stencil used for the
discretization. After $k$ transitions in the random walk, the history
will have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our our chord is of length $l$ and there are $n_i$ grid elements
(or states to which a history may transition) along that chord, then
$h = l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimension of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interpret $\tau$ to be the
\textit{effective optical thickness} of the domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

@@

To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix Eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix Eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest Eigenvalue of the diffusion operator
using an iterative Eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative Eigenvalue
solver.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4in,clip]{measured_spec_rad.png}
    \end{center}
    \caption{\textbf{Measured and analytic preconditioned diffusion
        operator spectral radius as a function of the absorption cross
        section to scattering cross section ratio.} \textit{Values of
        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
        computed numerically by an Eigensolver while the black dashed
        data was generated by Eq~(\ref{eq:iteration_radius}).}}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

@@

With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and the absorption cross varied from 0
to 100 while the scattering cross section was fixed at unity. Three
weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
used with 10,000 histories generated by a point source source of
strength 1 in the center of the domain such that the boundary
conditions will have less of an effect on the length of the random
walks. For each of the histories, the number of transitions made was
tallied to provide an effective value of $k$ for each history. This
value was then averaged over all histories to get a measured value of
$k$ for the particular operator. On the left,
figure~\ref{fig:measured_length} these measurements as well as the
analytic result computed by Eq~(\ref{eq:analytic_k}) as a function of
the iteration matrix spectral radius, $\rho(\ve{H})$. On the right,
figure~\ref{fig:measured_length} gives the relative error between the
predicted and observed results. We note good qualitative agreement
between the measured and analytic results. However, we observe a
larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=6.5in,clip]{measured_length.png}
    \end{center}
    \caption{\textbf{Measured and analytic random walk length as a
        function of the iteration matrix spectral radius.} \textit{The
        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
        \sn{1}{-8}. In the left plot, the red data was computed
        numerically by an adjoint Neumann-Ulam implementation while
        the black dashed data was generated by
        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
        error between the predicted and measured results is
        presented.}}
    \label{fig:measured_length}
  \end{spacing}
\end{figure}

@@

Finally, we seek to measure the leakage from a domain in a domain
decomposed Monte Carlo calculation and assess the quality of our
analytic relation for the optical thickness of domain and the
associated leakage approximations. For this experiment, a square grid
with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
direction with measurements occurring in the central domain without
boundary grid points. For the cross sections, the absorption cross was
varied from 1 to 100 while the scattering cross section was set to
zero to create a purely absorbing environment. The optical thickness
of these domains will vary as a function of the absorption cross
section if the other parameters are fixed. To compute the optical
thickness, along with the spectral radius as given by
Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
$n_s$ which respectively describe the typical domain length and the
average number of states moved along that typical length per history
transition. For our grid above, the domains are varied in size with
$50 \times 50$, $100 \times 100$, and $200 \times 200$ cells giving
$n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along the
typical length of the domain respectively with weight cutoff of
\sn{1}{-4}. Looking at the Laplacian stencil in
Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
will only move a single state in either the $i$ or $j$ directions due
to the symmetry of the problem. Furthermore, if we choose the $i$
direction, not all states we will transition to will move the history
in that direction. Therefore, we look to the definition of the
iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
to estimate the $n_s$ parameter. For a particular transition starting
at state $(i,j)$, 6 of the 8 possible new states in the stencil move
the history in $i$ direction with relative coefficients of 4 for
moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
i,\pm j)$. These coefficients dictate the frequency those states are
visited relative to the others. For those 6 states we can visit along
the typical length, their sum is 12 out of the total 20 for the
coefficients for all possible states with their ratio giving $n_i =
\frac{3}{5}$.

>>>>

%\usepackage{epsf}
%
%
% Insert authors' names and short version of title in lines below
%
\newcommand{\authorHead}      % Author's names here
   {Slattery, Wilson, and Evans}  
\newcommand{\shortTitle}      % Short title here
   {Domain Decomposed Neumann-Ulam Method Analysis}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%   BEGIN DOCUMENT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\section*{ABSTRACT} 
\begin{quote}
\begin{small}
The domain decomposed behavior of the adjoint Neumann-Ulam Monte Carlo
method for solving linear systems is analyzed using the spectral
properties of the linear operator. Relationships for the average
length of the adjoint random walks, a measure of convergence speed and
serial performance, are made with respect to the Eigenvalues of the
linear operator. In addition, relationships for the effective optical
thickness of a domain in the decomposition are presented based on the
spectral analysis and diffusion theory. Using the effective optical
thickness, the Wigner rational approximation and the mean chord
approximation are applied to estimate the leakage fraction of
stochastic histories from a domain in the decomposition as a measure
of parallel performance and potential communication costs. The
one-speed, two-dimensional neutron diffusion equation is used as a
model problem to test the models for symmetric operators. In general,
the derived approximations show good agreement with measured
computational results.

\emph{Key Words}: Monte Carlo, domain decomposition, MCSA, parallel
computing, linear solvers
\end{small} 
\end{quote}

@@

To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \cite{alexandrov_efficient_1998}. In order to enable higher
fidelity reactor physics Monte Carlo simulations, domain decomposition
has been identified as a key principle in moving forward in high
performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because these nearest-neighbor
communication sequences have poor algorithmic strong scaling
\cite{gropp_high-performance_2001}.

The purpose of this study is to provide a simple, analytic theory
based on the properties of the linear system that will allow for
estimates of the domain decomposed behavior of the adjoint
Neumann-Ulam method. When solving problems where the linear operator
is symmetric, a host of analytic theories exist based on the
Eigenvalue spectrum of the operator that characterize their behavior
in the context of deterministic linear solvers. Using past work, these
theories are adapted to the domain decomposed adjoint Neumann-Ulam
method using the one-speed, two-dimensional neutron diffusion
equation. In this paper we describe the adjoint Neumann-Ulam Monte
Carlo method followed by a presentation of the model problem. Using
the linear system generated by the discretization of this, we use a
spectral analysis to generate analytic relations for the Eigenvalues
of the operator based on system parameters. Using the Eigenvalue
spectra, we then build relationships to characterize the transport of
stochastic histories in a decomposed domain and the fraction of
histories that leak from a domain and will therefore have to be
communicated. Finally, we compare these analytic results to numerical
experiments conducted with the model problem and draw conclusions
looking towards future work.

@@

In order to perform the random walks required to approximate the
summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build the
column-scaled transition probability matrix and associated transition
weights as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
  \frac{h_{ji}}{p_{ij}},
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state, $j$, from the current
state, $i$, in the random walk by sampling the $i^{th}$ row of the
probablity matrix as a discrete distribution.  Using our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the Monte Carlo estimator for the solution in
state $i$ for a particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $W_m$ is the history weight on the $m^{th}$
step. Finally, the expectation value using all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
where $b_{i_0}$ will be the sampled source starting weight, therefore
also providing an unbiased Monte Carlo estimate of the solution.

@@

The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the Eigenvalues of the iteration matrix. We
will compute these Eigenvalues by assuming Eigenfunctions of the form
\cite{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
Eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these Eigenfunctions should
give the Eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

@@

The spectral radius of the iteration matrix is obtained by seeking its
largest Eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the Eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{equation}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}]\:.
  \label{eq:iteration_stencil}
\end{equation}
Inserting the Eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. We find that the maximum Eigenvalue exists when $p=q=0$,
giving the following for the spectral radius of the Jacobi
preconditioned iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

@@

In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a random
walk of average length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution})
\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
length, or the number of transitions before the termination of a
history (either by the weight cutoff, absorption, or exiting the
global domain) is therefore approximately the number of stationary
iterations required to converge to the specified tolerance. In the
case of the adjoint Neumann-Ulam method, no such tolerance exists,
however, we have specified a weight cutoff $W_c$ under which
low-weight histories will be prematurely terminated as their
contributions are deemed minute. After $k$ iterations, a stationary
method is terminated as the error has reached some fraction,
$\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
Neumann-Ulam method, if we take this fraction to be the weight cutoff,
a measure of how accurately the contributions of a particular history
to the solution are tallied, we then have the following relationship
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
Eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method.

@@

In the case of domain decomposed linear operator equations, we can use
diffusion theory to estimate the optical thickness of a domain in the
decomposition and the corresponding leakage fraction in terms of
properties of the linear operator and the discretization. To begin we
must first calculate the mean distance a Monte Carlo history will move
in the grid by computing the mean squared distance of its movement
along the chord of length $l$ defined across the domain. After a
single transition a history will have moved a mean squared distance
of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the discrete grid elements along the chord
and $n_s$ is the number of grid elements a history will move on
average every transition. For our diffusion model problem, $n_s$ would
equate to the number of states in the $i$ (or $j$ as the problem is
symmetric) direction that a history could potentially move in a single
transition and is dependent on the stencil used for the
discretization. After $k$ transitions in the random walk, the history
will have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our our chord is of length $l$ and there are $n_i$ grid elements
(or states to which a history may transition) along that chord, then
$h = l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimension of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo history in the domain. We can readily interpret $\tau$ to be the
\textit{effective optical thickness} of a domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

@@

To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix Eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix Eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest Eigenvalue of the diffusion operator
using an iterative Eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative Eigenvalue
solver.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4in,clip]{measured_spec_rad.png}
    \end{center}
    \caption{Measured and analytic preconditioned diffusion operator
      spectral radius as a function of the absorption cross section to
      scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
      $h=1.0$ were used. The red data was computed numerically by an
      Eigensolver while the black dashed data was generated by
      Eq~(\ref{eq:iteration_radius}).}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

@@

With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and the absorption cross varied from 0
to 100 while the scattering cross section was fixed at unity. Three
weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
used with 10,000 histories generated by a point source of strength 1
in the center of the domain. For each of the histories, the number of
transitions made was tallied to provide an effective value of $k$ for
each history. This value was then averaged over all histories to get a
measured value of $k$ for the particular operator. On the left,
Figure~\ref{fig:measured_length} presents these measurements as well
as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
the right, Figure~\ref{fig:measured_length} gives the relative error
between the predicted and observed results. We note good qualitative
agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=6.5in,clip]{measured_length.png}
    \end{center}
    \caption{Measured and analytic random walk length as a function of
      the iteration matrix spectral radius. The weight cutoff was
      varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
      plot, the red data was computed numerically by an adjoint
      Neumann-Ulam implementation while the black dashed data was
      generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
      relative error between the predicted and measured results is
      presented for each weight cutoff.}
    \label{fig:measured_length}
  \end{spacing}
\end{figure}

@@

Finally, we seek to measure the leakage from a domain in a domain
decomposed Monte Carlo calculation and assess the quality of our
analytic relation for the optical thickness of domain and the
associated leakage approximations. For this experiment, a square grid
with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
direction with measurements occurring in the central domain without
boundary grid points. For the cross sections, the absorption cross was
varied from 1 to 100 while the scattering cross section was set to
zero to create a purely absorbing environment with weight cutoff of
\sn{1}{-4}. The optical thickness of these domains will vary as a
function of the absorption cross section if the other parameters are
fixed. To compute the optical thickness, along with the spectral
radius as given by Eq~(\ref{eq:iteration_radius}), we also need the
parameters $n_i$ and $n_s$ which respectively describe the typical
domain length and the average number of states moved along that
typical length per history transition. For our grid above, the domains
are varied in size with $50 \times 50$, $100 \times 100$, and $200
\times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$ grid
points or states along the typical length of the domain
respectively. Looking at the Laplacian stencil in
Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
will only move a single state in either the $i$ or $j$ directions due
to the symmetry of the problem. Furthermore, if we choose the $i$
direction, not all states we will transition to will move the history
in that direction. Therefore, we look to the definition of the
iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
to estimate the $n_s$ parameter. For a particular transition starting
at state $(i,j)$, 6 of the 8 possible new states in the stencil move
the history in $i$ direction with relative coefficients of 4 for
moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
i,\pm j)$. These coefficients dictate the frequency those states are
visited relative to the others. For those 6 states we can visit along
the typical length, their sum is 12 out of the total 20 for the
coefficients for all possible states with their ratio giving $n_i =
\frac{3}{5}$.
